Effect of temporal sampling

### 2019-10-17 R/temporal-sampling.Rmd
adding the figure of error against initial number sampled

fig_cap <- captioner::captioner()
tab_cap <- captioner::captioner("Table")
```

@@

```{r read zurich data, include=T}
zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt"))
```

@@

```{r coniss, include=T, echo=F}
require(rioja)
diss<-dist(sqrt(zh[,-c(1)]/100)^2)
clust<-chclust(diss,method="coniss")
par(mfrow=c(2,1))
par(mar=c(5.1,4.1,2.1,20))
bstick(clust, cex=0.8)
par(mar=c(4,3.1,1.1,1.1))
plot(clust, labels=zh[,1], cex=0.8)
```
`r fig_cap("coniss clustering zurich", "Selection of significant groups and constrained hierarchical clustering of Lake Zürich sequence")`

```{r strat plot zurich, include=T, echo=F}
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(zh[,-c(1)],yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=colnames(zh[-c(1)]), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
```
`r fig_cap("strat plot zurich", "Stratigraphic abundance of cladoceran remains in Lake Zürich sediment core ZH17-21. (I'm aware it's impossible to read it with that many taxa, I have the same code somewhere applied to sub groups which make it way easier to read)")`

```{r strat plot zurich selected sp, include=T, echo=F}
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(cbind(zh[,colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")], rowSums(zh[,-1][,!colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")])),yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=c(colnames(zh[colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")]), "others"), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
```
`r fig_cap("strat plot zurich selected sp", "Stratigraphic abundance of dominant cladoceran remains in Lake Zürich sediment core ZH17-21. Abbreviations: BL= Bosmina longirostris, EC= Eubosmina coregoni, EL= Eubosmina longispina, DL= Daphnia spp., LK= Leptodora kindti, BYL= Bythotrephes longismanus, SC= Sida crystallina")`

@@

```{r eval=FALSE, include=FALSE}
require(vegan)
require(ade4)
require(changepoint)

# Change here the sample size
sample_size <- c(rep(5:40, each=3),50)

@@

line=0
for (i in sample_size) {
  line=line+1
  # 1. Sample years
  which_samples <- sample(1:nrow(zh), size = i,replace = F)
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

@@

```{r}
require(ggplot2)
require(gridExtra)
p1 <- ggplot(output, aes(Number_samples, PC1_var)) + geom_point() + stat_smooth()
p2 <- ggplot(output, aes(Number_samples, PC2_var)) + geom_point() + stat_smooth()
grid.arrange(p1,p2, nrow=1)
```
`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`

```{r}
p1 <- ggplot(output, aes(AMOC_PC1, Number_samples)) + geom_point() + xlim(1850,2017)
p2 <- ggplot(output, aes(AMOC_PC2,Number_samples)) + geom_point() + xlim(1850,2017)
grid.arrange(p1,p2, nrow=2)
```
`r fig_cap("timing of cpts", "Timing of changepoint")`

@@

```{r}
# Change here the sample size
sample_size <- c(rep(13:30, each=100))
#sample_size <- 15
output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")

@@

@@

  # Save outputs
  output$Initial_number_samples[i]     <- sample_size[i]
  output$Initial_AMOC_PC1[i]           <- cptmean.AMOC.PC1
  #output$Initial_AMOC_PC2[i]           <- cptmean.AMOC.PC2
  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1) <8){
        output$Number_needed_for_AMOC_PC1[i] <- length(which_samples)
        numberPC1_ok = TRUE
        } else numberPC1_ok = FALSE
  # if(!is.na(cptmean.AMOC.PC2) & !is.na(true.cptmean.AMOC.PC2) & cptmean.AMOC.PC2==true.cptmean.AMOC.PC2) {
  #   output$Number_needed_for_AMOC_PC2[i] <- length(which_samples) 
  #   numberPC2_ok = TRUE
  # } else numberPC2_ok = FALSE

@@

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point(alpha=.3)
library(dplyr)
output2 <- output %>% 
  group_by(Initial_number_samples) %>% 
  summarise(total0 = sum(total_added==0))

plot(output2)
abline(h=80)
```
`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`

@@

    # Save outputs
    c(nrow(zh), true.eig1, true.eig2,
      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
  } 
  # 1. Sample years
  which_samples <- sample(1:nrow(zh), size = sample_size[i],replace = F)
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

@@

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point()
```
`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`

@@

>>>>

fig_cap <- captioner::captioner()
tab_cap <- captioner::captioner("Table")

# Libraries
library(dplyr)
require(rioja)
require(vegan)
require(ade4)
require(changepoint)
require(ggplot2)
require(gridExtra)
require(plotly)
library(reshape)
```

@@

```{r read zurich data, include=T}
zh <- read.delim(paste0(getpath4data(),"ZH17_21_clado.txt")); is.zh=T
zh <- read.delim(paste0(getpath4data(),"VAR10-10-clado.txt")); is.zh=F
```

@@

```{r coniss, include=T, echo=F}
diss<-dist(sqrt(zh[,-c(1)]/100)^2)
clust<-chclust(diss,method="coniss")
par(mfrow=c(2,1))
par(mar=c(5.1,4.1,2.1,20))
bstick(clust, cex=0.8)
par(mar=c(4,3.1,1.1,1.1))
plot(clust, labels=zh[,1], cex=0.8)
```
<br>`r fig_cap("coniss clustering zurich", "Selection of significant groups and constrained hierarchical clustering of Lake Zürich sequence")`

```{r strat plot zurich, include=T, echo=F}
par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(zh[,-c(1)],yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=colnames(zh[-c(1)]), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
```
<br>`r fig_cap("strat plot zurich", "Stratigraphic abundance of cladoceran remains in Lake Zürich sediment core ZH17-21. (I'm aware it's impossible to read it with that many taxa, I have the same code somewhere applied to sub groups which make it way easier to read)")`

```{r strat plot zurich selected sp, include=T, echo=F}
if(is.zh){
  par(mfrow=c(1,1))
par(mar=c(5.1,4.1,2.1,20))
clado.clust<-strat.plot(cbind(zh[,colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")], rowSums(zh[,-1][,!colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")])),yvar=zh[,1],col.line="black",lwd.line=0,col.bar="black",lwd.bar=2, x.names=c(colnames(zh[colnames(zh) %in% c("LK","SC","DL","EL","EC","BL","BYL")]), "others"), cex.xlabel=1, srt.xlabel=30,scale.minmax=FALSE, y.rev = F,clust=clust)
addClustZone(clado.clust,clust,4,col="black")
}
```
<br>`r fig_cap("strat plot zurich selected sp", "Stratigraphic abundance of dominant cladoceran remains in Lake Zürich sediment core ZH17-21. Abbreviations: BL= Bosmina longirostris, EC= Eubosmina coregoni, EL= Eubosmina longispina, DL= Daphnia spp., LK= Leptodora kindti, BYL= Bythotrephes longismanus, SC= Sida crystallina")`

@@

```{r include=FALSE}
# Change here the sample size
sample_size <- c(rep(5:40, each=3),50)

@@

line=0
for (i in sample_size) {
  line=line+1
  # 1. Sample years
  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
  which_samples <- c(which_samples,1,nrow(zh))
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

@@

```{r}
p1 <- ggplot(output, aes(Number_samples, PC1_var)) + geom_point() + stat_smooth()
p2 <- ggplot(output, aes(Number_samples, PC2_var)) + geom_point() + stat_smooth()
grid.arrange(p1,p2, nrow=1)
```
<br>`r fig_cap("percentage explained by PCs", "Percentage explained by PC axes according to number of samples analysed")`

```{r}
p1 <- ggplot(output, aes(AMOC_PC1, Number_samples)) + geom_point() + xlim(1850,2017)
p2 <- ggplot(output, aes(AMOC_PC2,Number_samples)) + geom_point() + xlim(1850,2017)
grid.arrange(p1,p2, nrow=2)
```
<br>`r fig_cap("timing of cpts", "Timing of changepoint")`

@@

```{r}
# Change here the sample size
sample_size <- c(rep(13:30, each=30))
#sample_size <- 15
output <- as.data.frame(matrix(rep(NA, 5*length(sample_size)), ncol=5))
colnames(output) <- c("Initial_number_samples","Initial_AMOC_PC1","Initial_AMOC_PC2","Number_needed_for_AMOC_PC1","Number_needed_for_AMOC_PC2")

@@

    # # 4. Run changepoints on PC2
    true.cptmean.AMOC.PC2 <- NA
    # true.cptmean.AMOC.PC2 <- cpt.mean(PCAout$li[,2], method="AMOC") 
    # true.cptmean.AMOC.PC2 <- as.numeric(mdata[,1][cpts(true.cptmean.AMOC.PC2)])
    # if(length(true.cptmean.AMOC.PC2)==0) true.cptmean.AMOC.PC2 <- NA

@@

@@

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point(alpha=.3)

```
<br>`r fig_cap("how many samples must be added to get true changepoint", "A 'biased' sampling (trying to detect the true shift by getting closest sample) is not a guarantee to reduce the total number of samples that should be analyzed to detect the true shift")`

@@

    # Save outputs
    c(nrow(zh), true.eig1, true.eig2,
      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
  } 
  # 1. Sample years
  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
  which_samples <- c(which_samples,1,nrow(zh))
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

@@

```{r}
ggplot(output, aes(Initial_number_samples,total_added)) + geom_point()
```
<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`

```{r}
output2 <- output %>% 
  group_by(Initial_number_samples) %>% 
  summarise(total0 = sum(total_added==0))

plot(output2)
abline(h=80)
```

# Chances of getting close to the real break

Following discussion with Easton on 10/10/2019.

```{r}
# Change here the threshold year: how close you need to get close to the actual year
threshold_yr = c(2,4,6,8,10,12,14)

# Change here the sample size
sample_size <- c(rep(15:50, each=100))
#sample_size <- 15
output3 <- as.data.frame(matrix(rep(NA, 4*length(sample_size)*length(threshold_yr)), ncol=4))
colnames(output3) <- c("Initial_number_samples","threshold_yr","Initial_AMOC_PC1","Within_threshold")

for (i in 1:length(sample_size)) {
  if(i == 1)  {
    row=0
    mdata <- zh[order(zh[,1]),]
    # 2. Do ordination 
    PCAout<-decostand(mdata[,-1],"hellinger",na.rm=TRUE)
    PCAout<-dudi.pca(PCAout, scannf = FALSE)
    true.eig1 <- PCAout$eig[1]/sum(PCAout$eig)*100
    true.eig2 <- PCAout$eig[2]/sum(PCAout$eig)*100

    # Save outputs
    c(nrow(zh), true.eig1, true.eig2,
      true.cptmean.AMOC.PC1, true.cptmean.AMOC.PC2)
  }

  for (j in 1:length(threshold_yr)) {
    row=row+1
      # 1. Sample years
  which_samples <- sample(2:(nrow(zh)-1), size = sample_size[i],replace = F)
  which_samples <- c(which_samples,1,nrow(zh))
  which_samples <- which_samples[order(which_samples)]
  mdata <- zh[which_samples,]
  # reorder chronologically
  mdata <- mdata[order(mdata[,1]),]
  #remove rare species
  # present less than 3 times in the assemblage
  mdata <- mdata[,colSums(mdata>0)>3]

  # 3. Run changepoints on PC1
  cptmean.AMOC.PC1 <- cpt.mean(PCAout$li[,1], method="AMOC") 
  cptmean.AMOC.PC1_num <- cptmean.AMOC.PC1
  cptmean.AMOC.PC1 <- as.numeric(mdata[,1][cpts(cptmean.AMOC.PC1)])
  if(length(cptmean.AMOC.PC1)==0) cptmean.AMOC.PC1 <- 0

  # Save outputs
  output3$Initial_number_samples[row]     <- sample_size[i]
  output3$threshold_yr[row]               <- threshold_yr[j]
  output3$Initial_AMOC_PC1[row]           <- cptmean.AMOC.PC1
  if(abs(cptmean.AMOC.PC1-true.cptmean.AMOC.PC1)<=threshold_yr[j]){
        numberPC1_ok = 1
        } else {numberPC1_ok = 0}

  output3$Within_threshold[row] <- numberPC1_ok
  }

}
head(output3)
tail(output3)
```

```{r summary }
names(output3)

outsumm <- as.data.frame(with(output3, tapply(Within_threshold,list("Initial_number_samples"=Initial_number_samples,"threshold_yr"=threshold_yr), sum)))
outsumm$InitialNumber <- as.numeric(rownames(outsumm))
outsumm <- melt(outsumm, id.vars = "InitialNumber")
outsumm$more80 <- ifelse(outsumm$value>=80,1,0)
ggplot(outsumm, aes(InitialNumber, variable, fill=more80)) + geom_tile() +
  xlab("Sample analyzed") + ylab("Tolerance for error (year)")

```
<br>`r fig_cap("how many samples must be added to get true changepoint method 2", "Adding progressively samples to fill in the gaps between transition and next/previous sample is a better strategy.")`

@@

Is changepoint necesseraly what we want? We can be interested in changes in mean and/or variance. Right now, I've just looked at mean.
